Student Log-Data from a Randomized Evaluation of Educational Technology: A Causal Case Study

### 2019-10-28 paper.Rnw
obs. study & mediation sections

<<init,include=FALSE>>=
library(knitr)
opts_chunk$set(
echo=FALSE,results='asis',warning = FALSE,message = FALSE,cache=TRUE,error = FALSE
)

@

<<runStuff,include=FALSE,eval=FALSE>>=
source('R/dataAnalysis.r')
big <- sapply(ls(),function(x) object.size(get(x))>2*1e7)
save(list=ls()[!big],file='output/analysis.RData')
@

<<loadStuff,include=FALSE>>=
load('output/analysis.RData')
@

@@

All told, the main analysis included $n=$\Sexpr{prettyNum(sdat$nstud,',')} students, \Sexpr{prettyNum(sum(sdat$Z),',')} of whom were assigned to the CTA1 condition and \Sexpr{prettyNum(with(sdat,nstud-sum(Z)),',')} of whom were assigned to control.
The students were nested within \Sexpr{sdat[['nteacher']]} teachers, in \Sexpr{sdat[['nschool']]} schools across five states.

@@

The observational study and (especially) the mediation analysis we
describe below require a dichotomous measurement of hint usage: is a
student a high or low hint user?
To this end, we dichotomize $\bar{h}$ by comparing it to a cutoff value, so
that $H_i=\indicator{h_i>c}$, where $\indicator{x}=1$ if $x$ is true
and 0 otherwise.
To choose $c$, we fit a two-component modified Rasch mixture model
\citep{rasch} to hint request data, described in an online appendix.
The model classified approximately \Sexpr{round(mixProp*100)}\% of students as
high-hint users, so we chose $c=0.6$, approximately the
\Sexpr{round((1-mixProp)*100)}th percentile of $\bar{h}$.
This definition agreed with the model's classification in
approximately \Sexpr{round(agree*100)}\% of cases.

@@

We constructed an optimal full matching design \citep{hansen2004} to
condition on estimated propensity scores $\hat{\pi}$ using the
\texttt{optmatch} package in \texttt{R} \citep{optmatch}.
We matched $H=0$ subjects to $H=1$ subjects in such a way as
to minimize the overall distance between subjects in matched sets in
the logit-transformed propensity scores, $logit(\hat{\pi})$.
We constrained the match so that students could only be matched within
schools.
This was for two reasons: first, schools determine a number of factors
that in turn may impact posttest scores, including baseline
student characteristics, pedagogical styles, and CTA1 usage patterns
\citep[see][]{descriptivePaper}, and second, doing so facilitated the mediation
analysis in the following section.
Each matched set was allowed to contain any positive number of $H=1$
and any positive number of $H=0$ subjects, resulting in matches of
variable size and composition.
For instance, one matched set included 30 $H=1$ subjects matched to a
single $H=0$ subject, and another matched 21 $H=0$ subjects to a
single $H=1$ subject.

Figure \ref{fig:balance} in the online appendix shows that
matching largely eliminated mean differences in pre-treatment
covariates between $H=1$ and $H=0$ students.
The only notable exception is in pretest scores: $H=1$ students tended
to have slightly higher pretest scores than their matched comparisons
with $H=0$.
The p-value from an omnibus covariate balance test
\citep{covBal} is \Sexpr{prettyNum(round(balMod[['overall']]['matched','p.value'],2))}.

@@

% The classic approach to estimating effects from blocked randomization
% designs is a fixed-effects regression \citep{fisher1926arrangement},
% regressing outcomes on an indicator for treatment assignment and a set
% of indicators for matched sets.
% The results from this analysis are displayed on the row labeled
% ``ANCOVA'' in Table \ref{tab:matchResults}, with ``HC3'' standard
% errors and 95\% confidence intervals.
% They suggest that a greater-than-median proclivity to request hints
% \emph{hurts} student learning by a large amount, roughly equal to the
% entire treatment effect of CTA1 estimated in
% \citet{pane2014effectiveness}.
Table \ref{tab:matchResults} gives three treatment effect estimates
under matched ignorability.
The rows labeled ATE and TOT weight the matched sets to
estimate the average effect of $H$ over the entire treatment
group and over those members of the treatment group with $H=1$,
respectively.
The row labeled ANCOVA of Table \ref{tab:matchResults} gives the
coefficient on $H$ from a regression of Y on $H$ along with indicators for matched set,
racial categories, grade, ESL status, economic disadvantage, and sex,
along with a natural spline with five degrees of freedom for pretest (the standard error
and confidence interval use the ``HC3'' sandwich estimator,
\citealt{sandwichPackage}).
This regression adjusts for imbalances in these covariates within
matched sets \citep[c.f.][]{biasAdjust}; its estimate is similar to
the others.

\begin{table}
\centering
\begin{tabular}{crllcll}
&&Est.&SE&95\% CI&$\rho^2$&$T_Z$\\
\hline
\multirow{4}{*}{Estimate}%&ANOVA& -0.22 &0.04 &[-0.30,-0.13]&0&0\\
&ATE&-0.24&0.05&[-0.33,-0.14]&0&0\\
&TOT&-0.21&0.06&[-0.33,-0.10]&0&0\\
&ANCOVA&-0.22&0.04&[-0.31,-0.14]&0&0\\
\hline
\multirow{2}{*}{Sensitivity}&[Pretest]&-0.22&&[-0.55,0.10]&0.17&-13.14\\
&[Pretest/2]&-0.22&&[-0.40,-0.05]&0.08&-6.57\\
\hline
\end{tabular}
\caption{Estimates with and without regression adjustment and
  sensitivity analysis of the effect of hint usage.}
\label{tab:matchResults}
\end{table}

@@

We found that hint requests have a negative effect: students who tend
to request hints more often than the median score lower on the
posttest, after controlling for observed covariates.
When hint usage is expressed numerically, we find a monotonically
decreasing relationship between hint requests and posttest scores---as
far as we can tell, the optimal strategy is to never request a hint.
However, this observed effect may be due to confounding.
If, say, weaker students ask for hints more often and score lower
on the posttest regardless of hint usage, posttest scores will be
anticorrelated with hint requests.
This could even be the case after control for pretest scores and other
covariates, if these failed to capture all of the relevant dimensions
of student ability.
We found that an unobserved confounder that predicts both hint usage
and posttest scores just as strongly as the pretest could explain the
entire observed relationship.
The data, coupled with the existence of such a missing covariate, are
consistent with both positive and negative effects of hint requests.
On the other hand, covariates that are as important as pretests are
quite rare in practice---this confounding scenario might not be
plausible after all.
A less extreme missing confounder, predicting $Y$ and $H$ with half
the strength of the pretest, cannot explain the observed relationship,
and is only consistent with negative effects of hint requests.

@@

Since the potential outcome $Y(0,1)$ is undefined in the CTA1 hints
study, any expression (potentially) including $Y(0,1)$ is also
undefined; this includes $\delta(0)$ and $\xi(1)$.
Indeed, since hint behavior is unobserved in the control group
altogether, conventional approaches to mediation analysis do not
apply.
Further, were we to operationalize hint behavior as a continuous
variable, as may seem natural, mediation analysis in this case would
also be nearly impossible.
This is because students assigned to the control group request exactly
zero hints, whereas merely nine members of the treatment group
requested no hints over the course of the study, and these students
barely used the tutor at all.
That being the case, the data provide little to no information about
the distribution of potential outcomes when the software is used
but no hints are requested.

Our solution was to dichotomize hint requests with a median split, as
described above, essentially re-defining the causal question.
Half of the treatment group and all of the control group has
$H=0$---requesting hints at a rate less than the (treatment group)
median.
To make sense of the respective roles of $H$ and $\hcont$, the
continuous measure on which it is based, we may imagine a process in
which students are assigned $H$, and under that constraint choose a
value for $\hcont$.
In the control group all students are assigned $H=0$ and $\hcont=0$, but
in the treatment group half of students have $H=1$ and half have
$H=0$, and $\hcont$ varies within those groups.
That structure implies a ``treatment mediator interaction''
where the effect of the mediator on the outcome varies by treatment
status.
Of course, in our example treatment mediator interactions are not
identified, since $H=0$ and $\hcont=0$ in the control group, and very
few treatment students asked for 0 hints.
In any event, dichotomizing hint requests likely undersells the role
of hints, but allows a tractable mediation analysis.

@@

This estimate is a close cousin of the ``TOT'' estimate in Table
\ref{tab:matchResults} from the previous section.
Recall that the TOT effect is the average effect of $H$ on treatment
group subjects with $H=1$.
Both estimate contrast outcomes between $H=0$ and $H=1$, and both use
data from only the treatment group.
TOT is the the difference between the average outcome for students in
the treatment group with $H=1$, and the average outcome under the
counterfactual scenario in which they all had $H=0$.
That is,
\begin{equation*}
TOT=\EE[Y(1,1)|Z=1,H=1]-\EE[Y(1,0)|Z=1,H=1]
\end{equation*}
$\hat{\Delta}$ estimates the difference between the average outcome
for students in the entire treatment group, and the average outcome
for the treatment group were $H=0$ for all students.
That is
\begin{equation*}
\Delta=\EE[Y(1)|Z=1]-\EE[Y(1,0)|Z=1]
\end{equation*}
This can be re-written (omitting conditioning on $Z$):
\begin{align*}
\begin{split}
\Delta&=\EE[Y(1,1)|H=1]Pr(H=1)+\EE[Y(1,0)|H=0]Pr(H=0)\\
&\qquad -\EE[Y(1,0)|H=1]Pr(H=1)-\EE[Y(1,0)|H=0]Pr(H=0)
\end{split}
\\[2ex]
&=(\EE[Y(1,1)|H=1]-\EE[Y(1,0)|H=1])Pr(H=1)
\end{align*}
Since $Pr(H=1)=1/2$ ($H$ was defined with a median split),
$\hat{\Delta}=TOT/2$.
Estimated thusly, the indirect effect is negative: $-0.11$, with
a standard error of 0.03.
This analysis suggests that the availability of excessive hints
actually \emph{lowers} the treatment effect.
CTA1 works not because of hints, but despite them.

What would the effect be without hints?
To answer that question, we estimate ``direct'' effects,
$\Xi=\EE[\xi(1)]=\EE[Y(1,H(0))]-\EE[Y(0,H(0))]$.
Like indirect effects, this expression includes the average potential
outcome $\EE[Y(1,H(0))]$; unlike indirect effects, includes a term for
potential outcomes under the control condition.
Since the main contrast of $\Xi$ is due to treatment assignment,
construction of the estimator must account for the randomization
scheme.
This may be done with a regression estimator.
One option is to construct outcome $\tilde{Y}$ as, simply, $Y$ for
subjects in the control group and $\hat{Y}(1,0)$ for subjects in the
treatment group.
Then, regress it on a dummy variable for treatment group and dummy
variables for randomization block.
The clustering of subjects within schools can be accounted for in a
number of ways, such as including random intercepts for school or
computing cluster-robust standard error estimates.
Following the latter route, and using the \texttt{clubSandwich}
package in \texttt{R} \citep{clubsandwich}, we estimated a direct effect of $0.16\pm
0.24$.
Including covariates in the regression produces a direct effect of
$0.19\pm 0.21$.

>>>>

<<init,include=FALSE>>=
library(knitr)
opts_chunk$set(
echo=FALSE,results='asis',warning = FALSE,message = FALSE,cache=TRUE,error = FALSE
)
pn <- function(...) prettyNum(...,big.mark=',')
library(xtable)
@

<<runStuff,include=FALSE,eval=FALSE>>=
source('R/dataAnalysis.r')
@

<<loadStuff,cache=FALSE>>=
load('output/analysis.RData')
@

@@

All told, the main analysis included $n=$\Sexpr{prettyNum(Ntot,',')} students, \Sexpr{pn(NtFinal)} of whom were assigned to the CTA1 condition and \Sexpr{pn(Ntot-NtFinal)} of whom were assigned to control.
The students were nested within \Sexpr{pn(nteacher)} teachers, in \Sexpr{pn(nschool)} schools across five states.

@@

The observational study and (especially) the mediation analysis we
describe below require a dichotomous measurement of hint usage: is a
student a high or low hint user?
To this end, we dichotomize $\bar{h}$ by comparing it to a cutoff value, so
that $H_i=\indicator{h_i>c}$, where $\indicator{x}=1$ if $x$ is true
and 0 otherwise.
To choose $c$, we fit a two-component modified Rasch mixture model
\citep{rasch} to hint request data, described in an online appendix.
The model classified approximately \Sexpr{round(mixProp*100)}\% of students as
high-hint users, so we chose $c$ as the
\Sexpr{round((1-mixProp)*100)}th percentile of $\bar{h}$,
approximately \Sexpr{sprintf('%.1f',quantile(hintAvgStud[['propHintStud']],1-mixProp))}.
This definition agreed with the model's classification in
approximately \Sexpr{round(agree*100)}\% of cases.

@@

We constructed an optimal full matching design \citep{hansen2004} to
condition on estimated propensity scores $\hat{\pi}$ using the
\texttt{optmatch} package in \texttt{R} \citep{optmatch}.
We matched $H=0$ subjects to $H=1$ subjects in such a way as
to minimize the overall distance between subjects in matched sets in
the logit-transformed propensity scores, $logit(\hat{\pi})$.
We constrained the match so that students could only be matched within
schools.
This was for two reasons: first, schools determine a number of factors
that in turn may impact posttest scores, including baseline
student characteristics, pedagogical styles, and CTA1 usage patterns
\citep[see][]{descriptivePaper}, and second, doing so facilitated the mediation
analysis in the following section.
Each matched set was allowed to contain any positive number of $H=1$
and any positive number of $H=0$ subjects, resulting in matches of
variable size and composition.
For instance, one matched set included \Sexpr{max(sapply(levels(m3),function(ll) sum(optmatch:::getZfromMatch(m3)[m3==ll])))} $H=1$ subjects matched to a
single $H=0$ subject, and another matched \Sexpr{max(sapply(levels(m3),function(ll) sum(!optmatch:::getZfromMatch(m3)[m3==ll])))} $H=0$ subjects to a
single $H=1$ subject.

Figure \ref{fig:balance} and Table \ref{tab:balance} in the online appendix show that
matching largely eliminated mean differences in pre-treatment
covariates between $H=1$ and $H=0$ students.
The only notable exception is in pretest scores: $H=1$ students tended
to have slightly higher pretest scores than their matched comparisons
with $H=0$.
The p-value from an omnibus covariate balance test
\citep{covBal} is \Sexpr{prettyNum(round(balMod[['overall']]['matched','p.value'],2))}.

@@

% The classic approach to estimating effects from blocked randomization
% designs is a fixed-effects regression \citep{fisher1926arrangement},
% regressing outcomes on an indicator for treatment assignment and a set
% of indicators for matched sets.
% The results from this analysis are displayed on the row labeled
% ``ANCOVA'' in Table \ref{tab:matchResults}, with ``HC3'' standard
% errors and 95\% confidence intervals.
% They suggest that a greater-than-median proclivity to request hints
% \emph{hurts} student learning by a large amount, roughly equal to the
% entire treatment effect of CTA1 estimated in
% \citet{pane2014effectiveness}.
Table \ref{tab:matchResults} gives three treatment effect estimates
under matched ignorability.
The rows labeled ATE and TOT weight the matched sets to
estimate the average effect of $H$ over the entire treatment
group and over those members of the treatment group with $H=1$,
respectively.
The row labeled ANCOVA of Table \ref{tab:matchResults} gives the
coefficient on $H$ from a regression of Y on $H$ along with indicators for matched set,
racial categories, grade, ESL status, economic disadvantage, and sex,
along with a natural spline with five degrees of freedom for pretest (the standard error
and confidence interval using the ``HC3'' sandwich estimator,
\citealt{sandwichPackage}).
This regression adjusts for imbalances in these covariates within
matched sets \citep[c.f.][]{biasAdjust}; its estimate is similar to
the others.

<<obsStudyEst,results='asis',echo=FALSE>>=

matchTEtab <-
  cbind(
    data.frame(Effect=c('ANOVA','ATE','TOT','ANCOVA')),
    rbind(anovat,atet,tott,ancovat)[,c('Estimate','Std. Error')],
    CI=vapply(list(anovaci,ateci,totci,ancovaci),printci,'a'),
    `$T_Z$`=0,
    `$\\rho^2$`=0
  )

for(tt in c(ttt['xirt'],ttt['xirt']/2))
  matchTEtab <- add_case(matchTEtab,
    Effect=ifelse(tt==ttt['xirt'],"[Pretest]","[Pretest/2]"),
    Estimate=ancovat[1],
    CI=printci(interval(tt,rrr['xirt'],b=ancovat[1], se=ancovat[2],df=ancova$df)),
    `$\\rho^2$`=rrr['xirt'],
    `$T_Z$`=tt)

matchTEtab <- cbind(
  ` `=c('\\multirow{4}{*}{Estimate}',rep('',3),'\\multirow{2}{*}{Sensitivity}',''),
  matchTEtab
)

matchTEtab <- add_case(matchTEtab,` `='\\hline',.before=5)#,rep(NA,ncol(matchTEtab)-1)),matchTEtab[5:6,])

print(xtable(matchTEtab,
  align='rllrrrrr',
  caption='Estimates with and without regression adjustment and sensitivity analysis of the effect of hint usage.',
  label='tab:matchResults'),
  include.rownames=FALSE,sanitize.text.function =function(x) x)
@

% \begin{table}
% \centering
% \begin{tabular}{crllcll}
% &&Est.&SE&95\% CI&$\rho^2$&$T_Z$\\
% \hline
% \multirow{4}{*}{Estimate}%&ANOVA& -0.22 &0.04 &[-0.30,-0.13]&0&0\\
% &ATE&-0.24&0.05&[-0.33,-0.14]&0&0\\
% &TOT&-0.21&0.06&[-0.33,-0.10]&0&0\\
% &ANCOVA&-0.22&0.04&[-0.31,-0.14]&0&0\\
% \hline
% \multirow{2}{*}{Sensitivity}&[Pretest]&-0.22&&[-0.55,0.10]&0.17&-13.14\\
% &[Pretest/2]&-0.22&&[-0.40,-0.05]&0.08&-6.57\\
% \hline
% \end{tabular}
% \caption{Estimates with and without regression adjustment and
%   sensitivity analysis of the effect of hint usage.}
% \label{tab:matchResults}
% \end{table}

@@

We found that hint requests have a negative effect: students who
requested hints more often than the \Sexpr{round((1-mixProp)*100)}th
percentile score lower on the posttest, after controlling for observed covariates.
When hint usage is expressed numerically, we find a monotonically
decreasing relationship between hint requests and posttest scores---as
far as we can tell, the optimal strategy is to never request a hint.
However, this observed effect may be due to confounding.
If, say, weaker students ask for hints more often and score lower
on the posttest regardless of hint usage, posttest scores will be
anticorrelated with hint requests.
This could even be the case after controlling for pretest scores and other
covariates, if these failed to capture all of the relevant dimensions
of student ability.
We found that an unobserved confounder that predicts both hint usage
and posttest scores just as strongly as the pretest could explain the
entire observed relationship.
The data, coupled with the existence of such a missing covariate, are
consistent with both positive and negative effects of hint requests.
On the other hand, covariates that are as important as pretests are
quite rare in practice---this confounding scenario might not be
plausible after all.
A less extreme missing confounder, predicting $Y$ and $H$ with half
the strength of the pretest, cannot explain the observed relationship,
and is only consistent with negative effects of hint requests.

@@

Since the potential outcome $Y(0,1)$ is undefined in the CTA1 hints
study, any expression (potentially) including $Y(0,1)$ is also
undefined; this includes $\delta(0)$ and $\xi(1)$.
Indeed, since hint behavior is unobserved in the control group
altogether, conventional approaches to mediation analysis do not
apply.
Further, were we to operationalize hint behavior as a continuous
variable, as may seem natural, mediation analysis in this case would
also be nearly impossible.
This is because students assigned to the control group request exactly
zero hints, whereas merely
\Sexpr{english(sum(hintAvgStud[['propHintStud']]==0))}
members of the treatment group
requested no hints over the course of the study, and these students
barely used the tutor at all.
That being the case, the data provide little to no information about
the distribution of potential outcomes when the software is used
but no hints are requested.

Our solution was to dichotomize hint requests, as
described above, essentially re-defining the causal question.
\Sexpr{round((mixProp)*100)}\% of the treatment group and all of the control group has
$H=0$---requesting hints at a rate less than the
\Sexpr{round((1-mixProp)*100)}th percentile of the treatment group.
To make sense of the respective roles of $H$ and $\hcont$, the
continuous measure on which it is based, we may imagine a process in
which students are assigned $H$, and under that constraint choose a
value for $\hcont$.
In the control group all students are assigned $H=0$ and $\hcont=0$, but
in the treatment group half of students have $H=1$ and half have
$H=0$, and $\hcont$ varies within those groups.
That structure implies a ``treatment mediator interaction''
where the effect of the mediator on the outcome varies by treatment
status.
Of course, in our example treatment mediator interactions are not
identified, since $H=0$ and $\hcont=0$ in the control group, and very
few treatment students asked for 0 hints.
In any event, dichotomizing hint requests likely undersells the role
of hints, but allows a tractable mediation analysis.

@@

This estimate is a close cousin of the ``TOT'' estimate in Table
\ref{tab:matchResults} from the previous section.
Recall that the TOT effect is the average effect of $H$ on treatment
group subjects with $H=1$.
Both estimate contrast outcomes between $H=0$ and $H=1$, and both use
data from only the treatment group.
TOT is the the difference between the average outcome for students in
the treatment group with $H=1$, and the average outcome under the
counterfactual scenario in which they all had $H=0$.
That is,
\begin{equation*}
TOT=\EE[Y(1,1)|Z=1,H=1]-\EE[Y(1,0)|Z=1,H=1]
\end{equation*}
$\hat{\Delta}$ estimates the difference between the average outcome
for students in the entire treatment group, and the average outcome
for the treatment group were $H=0$ for all students.
That is
\begin{equation*}
\Delta=\EE[Y(1)|Z=1]-\EE[Y(1,0)|Z=1]
\end{equation*}
This can be re-written (omitting conditioning on $Z$):
\begin{align*}
\begin{split}
\Delta&=\EE[Y(1,1)|H=1]Pr(H=1)+\EE[Y(1,0)|H=0]Pr(H=0)\\
&\qquad -\EE[Y(1,0)|H=1]Pr(H=1)-\EE[Y(1,0)|H=0]Pr(H=0)
\end{split}
\\[2ex]
&=(\EE[Y(1,1)|H=1]-\EE[Y(1,0)|H=1])Pr(H=1)
\end{align*}
Since $Pr(H=1)=$\Sexpr{round(1-mixProp,2)},
$\hat{\Delta}=$\Sexpr{round(1-mixProp,2)}$TOT$.
Estimated thusly, the indirect effect is negative: \Sexpr{round(tott[1]*(1-mixProp),2)}, with
a standard error of \Sexpr{round(tott[2]*(1-mixProp),2)}.
This analysis suggests that the availability of excessive hints
actually \emph{lowers} the treatment effect.
CTA1 works not because of hints, but despite them.

What would the effect be without hints?
To answer that question, we estimate ``direct'' effects,
$\Xi=\EE[\xi(1)]=\EE[Y(1,H(0))]-\EE[Y(0,H(0))]$.
Like indirect effects, this expression includes the average potential
outcome $\EE[Y(1,H(0))]$; unlike indirect effects, includes a term for
potential outcomes under the control condition.
Since the main contrast of $\Xi$ is due to treatment assignment,
construction of the estimator must account for the randomization
scheme.
This may be done with a regression estimator.
One option is to construct outcome $\tilde{Y}$ as, simply, $Y$ for
subjects in the control group and $\hat{Y}(1,0)$ for subjects in the
treatment group.
Then, regress it on a dummy variable for treatment group and dummy
variables for randomization block.
The clustering of subjects within schools can be accounted for in a
number of ways, such as including random intercepts for school or
computing cluster-robust standard error estimates.
Following the latter route, and using the \texttt{clubSandwich}
package in \texttt{R} \citep{clubsandwich}, we estimated a direct
effect of
\Sexpr{citopm(dir1.1ci)}.
Including covariates in the regression produces a direct effect of
\Sexpr{citopm(dir2.1ci)}.
